Modelling potential environmental impacts of science activity in Antarctica

### 2024-11-29 paper/paper.qmd
put paper on website

## Introduction
Overview of Antarctic science since when and where, its intensity etc. Background on international treaties, etc.

@@

The functional form of some of these functions is shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions. 
The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
As has been noted elsewhere [see @Herzog2010;@Campbell2019] these parameter settings for the Tobler function are based on a loose fit with secondary data on hiking speeds presented in Imhof [-@Imhof1950, pp. 217-220], although these exact values are frequently applied in the literature.

@@

```{r}
#| label: fig-hiking-functions
#| fig-cap: Hiking functions.

@@

results <- tibble(Slope = -50:50/100,
                  Tobler = tobler(Slope), 
                  Gaussian = gaussian(Slope),
                  Lorentz = lorentz(Slope),
                  Rees = rees(Slope)) |>
  pivot_longer(cols = -1) |>
  rename(Function = name, `Speed, km/h` = value)

ggplot(results) +
  geom_line(aes(x = Slope, y = `Speed, km/h`, colour = Function, group = Function)) +
  scale_colour_discrete_c4a_cat("brewer.set1")
```

An additional example is offered by @Wood2023 who include the local gradient of the terrain (i.e. the maximum slope at each location, not the slope in the direction of movement) as a parameter in the function.

There is no generally accepted preferred form for hiking functions, although Tobler’s function has been the most widely used.

Note the strange origin of the popular exponential Tobler function (in a working paper, based on data in a figure in a relatively obscure Swiss book from 1950, and... according to Herzog 2010 a very poor fit to those data in any case...)

@@

## Methods and results 
### Hiking functions
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian. Other approaches rely on rule-of-thumb hiking time calculations such as...

>>>>

## Introduction
Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.

@@

The functional form of some of these functions is shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions.

@@

```{r}
#| label: fig-hiking-functions
#| fig-cap: Example hiking functions.

@@

results <- tibble(Slope = -50:50/100,
                  Exponential = tobler(Slope), 
                  Gaussian = gaussian(Slope),
                  Lorentz = lorentz(Slope),
                  Rees = rees(Slope)) |>
  pivot_longer(cols = -1) |>
  rename(Function = name, `Speed, km/h` = value)

ggplot(results) +
  geom_line(aes(x = Slope, y = `Speed, km/h`, 
            colour = Function, group = Function), linewidth = 1) +
  scale_colour_discrete_c4a_cat("brewer.set2") +
  scale_x_continuous(limits = c(-0.5, 0.5), expand = c(0, 0)) +
  theme(panel.border = element_rect(linewidth = 0.5, fill = NA))
```

The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
These parameter values for the exponential functional form give _Tobler's hiking function_ but, as has been noted elsewhere, [see @Herzog2010;@Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp. 217-220].
Nevertheless these parameter values are widely applied in the literature.

All these hiking function forms are somewhat _ad hoc_. 
They exhibit desirable, expected properties: a peak speed at a slope near zero, which we expect to be slightly negative (i.e., downhill), and continuously falling speeds as the slope deviates from the slope of peak speed.
However, there is no theoretical basis for the specific functional forms listed in @tbl-hiking-function-equations.
More principled approaches might be developed based on the literature on the physiology of human movement, see e.g. **CITATIONS NEEDED HERE**.
In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
However, these approaches are hard to implement, and empirically-derived and locally-specific hiking functions as reported by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019 are the approach we have taken in the present work, based on available data and the goals of our study where the relative cost of different potential routes in the landscape is more important than exact prediction of routes.
In practice, almost any function with the peaked form of those shown in @fig-hiking-functions is suited to our requirements.

It is commonplace in many applications to also incorporate a penalty on movement contingent on landcover, especially for off-track or off-road movement.
For example the speed attainable off-track in forested terrain might be only half that attainable in grasslands.
Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020.
The hiking function reported by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function. 
It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be applied in future work. 
In our application because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover and estimate movement costs conditional on the landcover present at each location.
This has the additional advantage of allowing for distinctly different effects of slope on speed.
Details of this approach are reported and discussed in @sec-hiking-function-estimation.

@@

## Methods and results 
### Hiking functions {#sec-hiking-function-estimation}
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian. Other approaches rely on rule-of-thumb hiking time calculations such as...
